---
title: "Car Accidents and Population Growth in Baton Rouge, LA, Tempe, AZ, and Seattle, WA"
author: "Alejandro Calderon, Dylan Craker, Trent Douglas"
date: "5/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(readxl)
library(lubridate)
library(grDevices)
library(ggplot2)
```

```{r load, include=FALSE}
baton_rouge_data <- read.csv("data/Baton_Rouge_Traffic_Incidents.csv")
baton_rouge_population <- read_xlsx("data/baton_rouge_pop2.xlsx")
seattle_data <- read.csv("data/seattle_Collisions.csv")
seattle_population <- read_xlsx("data/seattle_population.xlsx")
tempe_data <- read.csv("data/tempe_Crash_Data_Report__detail_.csv")
national_data <- read_xlsx("data/nhtsa_yearly_national_stats.xlsx")
```
## 1. Introduction
For our final report we decided to study car crash data in 3 US cities, Baton Rouge, Louisiana; Seattle, Washington; and Tempe, Arizona. We chose these three cities because of their similarities and differences. Baton Rouge is a relatively stagnant city with respect to population and area, neither having grown much in the past 10 years, with population actually slightly declining. Seattle, because it has not increased in area, but its a;ready high population has increased at the third fastest rate in the country. Finally, we chose Tempe because has also been increasing a rapid rate, but it has a lower population than Seattle. The differing populations in these cities and their trends are what inclined us to chose them, as well as having the most available and most well documented data that we could find. Our research question in these three cities was: how accurate of a predictor is population density with respect to fatal car accidents?

Our hypothesis is that fatalities increase as population density increases. We believe this is the case, because the growth of a city is often correlated with increased economic activity, bringing more upper and middle class people into the city. This results in more cars in the city, and more drivers on the road. This can also create more foot traffic in areas of highly populated cities and result in more accidents, and therefore more fatal accidents. This is our hypothesis, we believe it is important to research this because it could help provide legislators and government officials with the information necessary to prevent excessive building permits, regulate road construction, and create overall better systems for traffic and prevent fatal accidents if possible. The opposite of our hypothesis would be that population density is not an accurate predictor of fatal car accidents.

## 2. Data Descriptions
We found out data of data.gov, a site that compiles statistical data that the government produces. There were tons of different datasets to chose from, whether national, state, city, town, county, etc. However, after looking through all the options, the Baton Rouge, Seattle, and Tempe datasets, seemed more complete, most detailed, and most up to date. Once we started looking at the data, we decided to narrow down our focus to more specific variables such as year, fatality, injury, and factors/causes of accidents.

## Baton Rouge Data
```{r baton_rouge_maniputalion, include=FALSE}
baton_rouge <- select(baton_rouge_data, CRASH.DATE, OCCURED.ON, FATALITY, INJURY, PRIMARY.FACTOR, SECOND.FACTOR, PEDESTRIAN)
br_dataframe <- sapply(baton_rouge, is.factor)
baton_rouge[br_dataframe] <- lapply(baton_rouge[br_dataframe], function(FATALITY) as.numeric(as.character(FATALITY)))
transform(baton_rouge, FATALITY =as.numeric(FATALITY))

baton_rouge$year <- strptime(as.character(baton_rouge$CRASH.DATE), "%m/%d/%Y")
baton_rouge$year <- format(baton_rouge$year, "%Y")
br_data <- select(baton_rouge, OCCURED.ON, FATALITY, INJURY, PRIMARY.FACTOR, SECOND.FACTOR, PEDESTRIAN, year)

```

```{r baton_rouge_data_cleaning, include=FALSE}
br_data$FATALITY[br_data$FATALITY == ""] <- 0
br_data$FATALITY[br_data$FATALITY == "X"] <- 1
br_data$FATALITY <- as.numeric(br_data$FATALITY)

br_data$INJURY[br_data$INJURY == ""] <- as.numeric(0)
br_data$INJURY[br_data$INJURY == "X"] <- as.numeric(1)
br_data$INJURY <- as.numeric(br_data$INJURY)

br_data$PEDESTRIAN[br_data$PEDESTRIAN == ""] <- as.numeric(0)
br_data$PEDESTRIAN[br_data$PEDESTRIAN == "X"] <- as.numeric(1)
br_data$PEDESTRIAN <- as.numeric(br_data$PEDESTRIAN)

br_data$SECOND.FACTOR[br_data$SECOND.FACTOR == ""] <- "NONE"
```

```{r br_summary}
br_summary <- summary(br_data)
br_summary
```

```{r br_desc_stats, include=FALSE}
primary_factors_br <- unique(br_data$PRIMARY.FACTOR)

accident_totals <- br_data %>% group_by("primary_reason" = br_data$PRIMARY.FACTOR) %>% count(name = "total_obs")
year_totals <- br_data %>% group_by("year" = br_data$year) %>% count(name = "year_totals")


br_fatality <- subset(br_data, subset = (FATALITY == 1)) 
fatality_counts <- br_data %>% group_by(year) %>% filter(FATALITY == 1) %>% count(name = "fatality_counts")
fatality_counts$year <- as.numeric(fatality_counts$year)

br_no_fatality <- subset(br_data, subset = (FATALITY == 0))
no_fatality_count <- br_no_fatality %>% group_by(year) %>% filter(FATALITY == 0) %>% count(name = "no_fatality_counts") %>% filter(year < 2021)
no_fatality_count$year <- as.numeric(no_fatality_count$year)
```

```{r fatality_plot_work, include=FALSE}
fatality_plot <- ggplot() +
  geom_point(data = fatality_counts, aes(year, fatality_counts), colour = "red")+
  geom_point(data = no_fatality_count, aes(year, no_fatality_counts), colour = "blue")+
  xlim(c(2010, 2020)) +
  scale_y_continuous(trans = 'log2') +
  ylab("Total Number of Fatalities") +
  xlab("Year")
```

```{r fatality_plot}
print(fatality_plot + ggtitle("Number of Car Accidents in Baton Rouge from 2010-2021"))
```

```{r descriptive, include=FALSE}
fatality_count_br <- br_fatality %>% group_by(br_fatality$year) %>% count()
no_fatality_count_br <- br_no_fatality %>% group_by(br_no_fatality$year) %>% count()
```
```{r descript_plot}
br_desc_plot <- ggplot(accident_totals, aes(x=total_obs, y=primary_reason)) +
  geom_bar(stat="identity") +
  ylab("Cause of Accident") +
  xlab("Frequency") +
  scale_x_continuous(trans = 'log2')
print(br_desc_plot + ggtitle("Cause of Car Accidents in Baton Rouge from 2010-2021"))
```

```{r summary stats, include=FALSE}
summary(accident_totals$total_obs)
```

```{r br_population, include=FALSE}
br_pop <- baton_rouge_population %>% filter(`Geographic Area` == "Baton Rouge city, Louisiana") %>% select("2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018", "2019")
br_pop_rotated <- br_pop %>% pivot_longer(br_pop, cols = starts_with("20"), names_to = "year", values_to = "pop")
baton_rouge_fat_no_fat <- merge(fatality_counts, no_fatality_count, by = "year")
```
```{r merged_data_baton_rouge}
baton_rouge_totals <- merge(baton_rouge_fat_no_fat, br_pop_rotated, by = "year")
br_area <- 88.65
baton_rouge_totals$pop_density <- (baton_rouge_totals$pop/br_area)
```

```{r br_regression}
baton_rouge_regression <- summary(lm(fatality_counts ~ pop_density + year, data = baton_rouge_totals))
baton_rouge_regression
```
The Baton Rouge R squared is too small to be described as accurately predicting the data. The p-value is too large to describe the correlation between car accident fatality and population changes from 2010-2019, seeing as it is greater than .05. From the size of the p-value, we fail to reject the null hypothesis with the Baton Rouge data set. 


```{r seattle}
##Filter unnecessary columns
seattle_data <- select(seattle_data, ADDRTYPE, INJURIES, SERIOUSINJURIES, FATALITIES, INCDTTM, SEVERITYDESC, UNDERINFL, SPEEDING)

##Change date to only year
seattle_data$Year <- strptime(as.character(seattle_data$INCDTTM), "%m/%d/%Y")
seattle_data$Year <- format(seattle_data$Year, "%Y")

##Change blanks to NA & fix columns
seattle_data$ADDRTYPE[seattle_data$ADDRTYPE == ""] <- NA
seattle_data$UNDERINFL[seattle_data$UNDERINFL == ""] <- NA
seattle_data$UNDERINFL[seattle_data$UNDERINFL == "0"] <- NA
seattle_data$SPEEDING[seattle_data$SPEEDING == ""] <- NA
seattle_data$SEVERITYDESC[seattle_data$SEVERITYDESC == "Unknown"] <- "Inconclusive"
seattle_data$FATALITIES[seattle_data$FATALITIES == "2"] <- "1"
seattle_data$FATALITIES[seattle_data$FATALITIES == "5"] <- "1"


##Subset to years in used in analysis
sea_data <- subset(seattle_data, subset = (Year >= "2012" & Year <= "2019"))
sea_data <- na.omit(sea_data)
sea_data_summary <- summary(sea_data)
sea_data_summary

accident_totals_sea <- sea_data %>% group_by("SEVERITYDESC" = sea_data$SEVERITYDESC) %>% count(name = "total_obs")
year_totals_sea <- sea_data %>% group_by("Year" = sea_data$Year) %>% count(name = "year_totals")
accident_totals_sea
year_totals_sea

sea_fatality <- subset(sea_data, subset = (FATALITIES == 1))
fatality_counts_sea <- sea_data %>% group_by(Year) %>% filter(FATALITIES == 1) %>% count(name = "fatality_counts")
fatality_counts_sea$Year <- as.numeric(fatality_counts_sea$Year)

sea_no_fatality <- subset(sea_data, subset = (FATALITIES == 0))
no_fatality_count_sea <- sea_no_fatality %>% group_by(Year) %>% filter(FATALITIES == 0) %>% count(name = "no_fatality_counts") %>% filter(Year < 2021)
no_fatality_count_sea$Year <- as.numeric(no_fatality_count_sea$Year)

# fatality_lm <- lm(fatality_counts_sea~Year, data = fatality_counts_sea)
# predicted_fatality <- data.frame(fatality_pred = predict(fatality_counts_sea, fatality_lm))

```

```{r}
##Did my best, not sure why its not running
fatality_plot_sea <- ggplot() +
  geom_point(data = fatality_counts_sea, aes(Year, fatality_counts), colour = "red")+
  # geom_line(colour = "red2", data = predicted_fatality, aes(x = Year, y = fatality_counts_sea)) +
  geom_point(data = no_fatality_count_sea, aes(Year, no_fatality_counts), colour = "blue")+
  # geom_smooth(method = 'lm', formula = no_fatality_count_sea~Year) +
  xlim(c(2012, 2019)) +
  scale_y_continuous(trans = 'log2') +
  ylab("Total Number of Fatalities") +
  xlab("Year")
print(fatality_plot_sea + ggtitle("Number of Car Accidents in Seattle from 2012-2019"))
```


```{r}
##Plot Severity of Accidents
sea_severity <- ggplot(sea_data) +
  geom_bar(aes(x = SEVERITYDESC)) +
  theme(axis.text.x = element_text(size = 6))+
  scale_y_continuous(trans = 'log2')+
  xlab("Accident Result")+
  ylab("Total Observed")
print(sea_severity + ggtitle("Total Severity of Accidents in Seattle"))
```
```{r}
seattle_fat_no_fat<- merge(fatality_counts_sea, no_fatality_count_sea, by = "Year")
seattle_totals <- merge(seattle_fat_no_fat, seattle_population, by = "Year")
seattle_area <- 83.78
seattle_totals$pop_density <- (seattle_totals$Population/seattle_area)

seattle_regression <- summary(lm(fatality_counts ~ pop_density + Year, data = seattle_totals))
seattle_regression
```

```{r tempe data manipulation, include=FALSE}
#initial filter of columns
tempe_filter <- tempe_data %>%
  select(Year, Totalinjuries, Totalfatalities, Injuryseverity, AlcoholUse_Drv1,
         DrugUse_Drv1, Unitaction_One, Violation1_Drv1, StreetName, CrossStreet, Lightcondition,
         JunctionRelation,Weather, SurfaceCondition) 

#filters out blank rows
tempe_filter <- tempe_filter[!(tempe_filter$AlcoholUse_Drv1 == ""), ]
```

```{r tempe summary}
summary(tempe_filter)
```

```{r tempe injury plot}
#simple barplot
tempe_severity_plot <- ggplot(tempe_filter) +
  geom_bar(aes(x = Injuryseverity)) +
  theme(axis.text.x = element_text(size = 6)) +
  xlab("Severity of Injury") +
  ylab("Number of Injuries")+
  scale_y_continuous(trans = 'log2')
print(tempe_severity_plot + ggtitle("Basic Plot Summarizing Injuries in Tempe"))
```
```{r tempe stacked bar plot}
#testing aggregating counts for stacked bar chart
tempe_aggregate <- count(tempe_filter, Year, Totalinjuries, AlcoholUse_Drv1)



tempe_ord <- mutate(tempe_aggregate,
                     Totalinjuries = reorder(Totalinjuries, -n, sum),
                     AlcoholUse_Drv1 = reorder(AlcoholUse_Drv1, -n, sum))


#stacked bar plot showing number of injuries in accordance to alcohol influence
tempe_alcinf <- ggplot(tempe_aggregate) +
  geom_col(aes(x = AlcoholUse_Drv1, y = n, fill = Totalinjuries))+
  scale_y_continuous(trans = 'log2') +
  xlab("Alcohol Use or No Apparent Use") +
  ylab("Scaled Injuries from Accidents")

#makes multiple charts for each year
print(tempe_alcinf + facet_wrap(~ Year) + ggtitle("Faceted Stacked Charts Showing Injuries Caused by Alcohol Influence"))

```

```{r tempe population, include=FALSE}
#tempe populations for future reference
tempepop_2012 <- 164139
tempepop_2013 <- 164742
tempepop_2014 <- 166975
tempepop_2015 <- 169816
tempepop_2016 <- 174708
tempepop_2017 <- 178339
tempepop_2018 <- 183652
tempepop_2019 <- 187454

#tempe area mi^2
tempe_area <- 40.22


#filters fatalities by year
tempe_population <- tempe_filter %>%
  group_by(Year) %>%
  summarise(Fatalities = sum(Totalfatalities)) %>%
  select(Year, Fatalities)

#filters accidents with no fatalites by year
tempe_no_fatal <- tempe_filter %>%
  group_by(Year) %>%
  filter(Totalfatalities == 0) %>%
  count(name = "no_fatalities")

#merges fatalities and no fatalities by year
tempe_totals <- merge(tempe_population, tempe_no_fatal)

tempe_totals$Year <- as.numeric(tempe_totals$Year)

#populations by year
tempe_pops <- c(tempepop_2012, tempepop_2013, tempepop_2014, tempepop_2015,
                 tempepop_2016, tempepop_2017, tempepop_2018, tempepop_2019)
#appends populations
tempe_totals$Population <- tempe_pops
#appends population density
tempe_totals$pop_density <- (tempe_totals$Population/tempe_area)

```
```{r tempe regression}
tempe_regression <- summary(lm(Fatalities ~ pop_density + Year, data = tempe_totals))
tempe_regression
```